Management induced changes of soil organic carbon on global croplands

### 2022-07-20 SOCbudgetPaper/SOCbudget.tex
update to version before bug check

\begin{abstract}
Soil organic carbon (SOC) is one of the largest terrestrial carbon (C) stocks on Earth. The first meter of the Earth's soils profile stores three times as much C as the vegetation and twice the amount of C in the atmosphere. SOC has been depleted by anthropogenic land-cover change and agricultural management. However, the latter has so far not been well represented in global C stock assessments. While SOC models often simulate detailed biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated at the global scale. Here we develop a spatially explicit data set for agricultural management on cropland, considering crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We combine it with a reduced-complexity model based on the IPCC Tier 2 method to create a half-degree resolution data set of SOC stocks and SOC stock changes for the first 30 cm of mineral soils. We estimate that due to arable farming, soils have lost around \(34.6\,\unit{GtC}\) relative to a counterfactual hypothetical natural state in 1975. Within the period 1975--2010 this SOC debt continued to expand by \(5\,\unit{GtC}\) (\(0.14\,\unit{GtC\,yr^{-1}}\)) to around 39.6 GtC. However, accounting for historical management led to \(2.1\,\unit{GtC}\) less (\(0.06\,\unit{GtC\,yr^{-1}}\)) emissions than under the assumption of constant management. We also find that management decisions have influenced the historical SOC trajectory most strongly by residue returning, indicating that increasing SOC sequestration by biomass retention may be a promising negative emissions technique. The reduced-complexity SOC model may allow to simulate management-induced SOC sequestration also within computationally demanding integrated (land-use) assessment modeling.
\end{abstract}

@@

Soil organic carbon (SOC), the amount of organic carbon stored in the Earth's soil, constitutes the largest terrestrial organic carbon pool. It exceeds the carbon in the atmospheric and vegetation pools multiple times \citep{batjes_total_1996}. Even small changes in processes affecting SOC lead therefore to substantial shifts in the terrestrial carbon cycle and influence the amount of CO\textsubscript{2} in the atmosphere \citep{frielingstein_glob_carbon_2020, minasny_soil_2017}. The specific amount of carbon stored in soils globally is quantified with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile \citep{batjes_total_1996, sanderman_soil_2017}.

@@

The anthropogenic impact can be measured by the SOC debt (also referred to as SOC component of land-use change emissions, see Pongratz et al.~\citeyearpar{pongratz_terminolgy_2014}), which is the amount of organic carbon soils have lost under cultivation compared to a hypothetical potential natural vegetation state. Sanderman et al.~\citeyearpar{sanderman_soil_2017} identified the anthropogenic SOC debt for the first meter of the soil profile due to land cover change at around 116 GtC (37 GtC for the first 30 cm), compared to previous estimates of 60--130 GtC for the first meter \citep{lal_world_2001}.

Global assessments of the carbon cycle via dynamic global vegetation models (DGVMs), Earth System Models (ESMs) or bookkeeping models (BKMs) have analyzed SOC losses as part of a comprehensive evaluation of the global carbon budget and land-use change (LUC) emissions \citep{frielingstein_glob_carbon_2020}. While providing estimates of the magnitude of SOC losses due to land-cover change, most models lack a detailed representation of agricultural management. Earlier DGVM- and ESM-based assessments only considered changes in land cover, but ignored the removal of biomass at harvest \citep{strassmann_simulating_2008, betts_climate_2015}. BKMs are designed to estimate LUC related emissions but often ignore changes in SOC due to climate change, CO\textsubscript{2} fertilization and N deposition. Whereas BKM have largely improved in estimating additional emissions from wood harvest and shifting cultivation, state of the art models do not consider impacts of varying agricultural management \citep{frielingstein_glob_carbon_2020, houghton_carbon_2012, hansis_relevance_2015, bastos_comparison_2021}.

Managed agricultural systems were introduced in greater detail to DGVMs and ESMs to improve the assessment of the terrestrial carbon balance \citep[e.g.~][]{bondeau_modelling_2007, lindeskog_implications_2013}. Pugh et al.~\citeyearpar{pugh_simulated_2015} explicitly consider agricultural management in the form of tillage, irrigation and biomass extraction at harvest, but worked with stylized scenarios rather than with historical management data. They also showed the importance of accounting for the land-use history, as many carbon emissions from agricultural soils are caused by historical LUC and the slow decline of SOC under cropland before a new equilibrium is reached.

In global-scale carbon cycle assessments, management systems are typically represented as spatially explicit patterns that are static in time (e.g.~growing seasons \citep{portmann_mirca2000global_2010}, multiple cropping systems \citep{waha_multiple_2020}, irrigation systems \citep{jagermeyr_water_2015}) or as stylized scenarios \citep[e.g.~][]{pugh_simulated_2015, lutz_simulating_2019}. Herzfeld et al.~\citeyearpar{herzfeld_soc_2021} account for historical changes in fertilizer and manure inputs, residue removal rates and tillage systems and report SOC losses from cropland expansion over the period from 1700--2018 of 215 GtC. Within their stylized future management scenarios they find that none of the management aspects considered (residue management, no-tillage) can create a net carbon sink on current cropland areas under future climate change.

More data sets on spatially explicit agricultural management time series with global coverage are becoming available (e.g.~on tillage systems, see \citep{porwollik_generating_2019, prestele_spatially_2018}) and modeling approaches are increasingly being developed to project the dynamics of management systems into the future (e.g.~\citep{iizumi_modeling_2019, minoli_modelling_2019}), but have --- to our knowledge --- not yet found their way into comprehensive assessments of the terrestrial carbon cycle in DGVMs and BKMs.

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks \(SOC^{\mathrm{eq}}\) are given by
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
\label{eq:inoutflow}
\end{equation}
with \(C^{\textrm{in}}\) being the carbon inputs to the soil, \(k\) denotes the soil organic carbon decay rate. This equation is valid for all grid cells \(i\) and all years \(t\). We use the Tier 2 modeling approach for our calculations, which assumes three soil carbon sub-pools \(sub\) (active, slow and passive) and interactions between them, following the approach in the Century model \citep{parton_analysis_1987}. Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type \(lu\) specific.
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and pastures (referred to as natural vegetation in the following).

@@

Whereas residue and manure default lignin and nitrogen fractions are given by the IPCC guidelines \citep{ogle_cropland_in_ipcc_2019}, we use plant-functional type and plant-organ specific parameterization for the natural litterfall. Global distribution of plant-functional types is given by \citep{schaphoff_lpjml4_2018} as well as separation of litter into leaf, fine root and wood litter compartments excluding litter biomass burnt in wild fires. Leaf litter parameters are given by Brovkin et al.~\citeyearpar{brovkin_plant-driven_2012}, fine root to leaf litter lignin ratio by Guo et al.~\citeyearpar{guo_leaf_root_decompostion_2021}, lignin content of wood litter by Rahman et al.~\citeyearpar{rahman_lignin_2013} and nitrogen content scaling factors for leaf to fine roots and leaf to wood litter by von Bloh et al.~\citeyearpar{vBloh_lpjml5_2018}.
Data sources for all considered carbon inputs as well as for lignin and nitrogen parameterization are listed in Table \ref{tab:datasourceinputs}.

@@

For natural vegetation, we assume rainfed and non-tilled conditions, whereas for cropland, we distinguish the effect of different tillage (see Sect. \ref{sec:tillage}) and irrigation (see Sect. \ref{sec:irrigation}) practices on decay rates. We calculate area-weighted means for \(till\) and \(wat\) on cropland for each grid cell, using area shares for the different tillage and irrigation practices. Data sources as well as used parameters for the different decay drivers for all land-use types are listed in Table \ref{tab:datasourcedecay}; equations are displayed by equations 5.0B--5.0F in Ogle et al.~\citeyearpar{ogle_cropland_in_ipcc_2019}.

@@

Firstly, in order to account for the impacts of legacy fluxes from land-use changes long before the time horizon of interest, we consider land-use change from 1510 onwards. In 1510, we assume all \(SOC\) pools to be in natural steady-state, implying that all land-use change prior to that time occurs in 1510. We assume that by 1901, all cropland converted in 1510 has reached its new steady state, so that it is not necessary to explicitly account for even older land conversion. Model inputs for 1901--1930 for climate and natural vegetation litterfall are repeated for 1510--1900 to mimic constant climate conditions for this first initializing period. Similarly, agricultural management data are held constant at the level of 1965 until 1965. This approach follows others studies looking on effects of land-use change and management \citep[e.g.~][]{schaphoff_lpjml4_2018-1, herzfeld_soc_2021}.

@@

Additionally to the Tier 2 modeling approach \citep{ogle_cropland_in_ipcc_2019} and the detailed analysis of management data coming with it, \(SOC\) changes can be estimated using the IPCC Tier 1 approach of IPCC guidelines \citep{eggleston_ipcc_2006, calvo_buendia_ipcc_2019}. Here, stocks are calculated via stock change factors (\(F^{\mathrm{SCF}}\)) given by the IPCC for the topsoil (0-30 cm) and based on observational data. Estimates of \(F^{\mathrm{SCF}}\) are differentiated by crops, management and input systems (here summarized under \(m\)) reflecting different dynamics under changed in- and outflows without explicitly tracking these flows. Moreover, estimates of \(F^{\mathrm{SCF}}\) vary for different climatic zones (\(c\)) specified by the IPCC (see Fig. \ref{fig:CLIMzone}). The actual \(SOC\) stocks are thus calculated based on a given reference stock \(SOC^{\mathrm{ref}}\) by

@@

with \(T_{c,i}\) being the translation matrix for grid cells \(i\) into corresponding climate zones \(c\). For this analysis, we use the default \(F^{\mathrm{SCF}}\) from the Tier 1 method of Eggleston et al.~\citeyearpar{eggleston_ipcc_2006} and Calvo Buendia et al.~\citeyearpar{calvo_buendia_ipcc_2019} as a comparison and consistency check for our more detailed Tier 2 steady-state approach.

@@

Crop residue production and management is based on a revised methodology of Bodirsky et al.~\citeyearpar{bodirsky_n2o_2012} and key aspects are explained as they play a central role in soil carbon modeling. Starting from crop yield estimates \(Y\) and respective physical crop area \(CA\), we estimate total above-ground \(AGR\) and below-ground \(BGR\) residue biomass (in tonnes) using crop group \(cg\) specific ratios for above-ground residues to harvested biomass \(r^{\mathrm{ag,prod}}_{cg}\) in \((tDM\,ha^{-1})(tDM\,ha^{-1})^{-1}\) above-ground residues to harvested area \(r^{\mathrm{ag,area}}_{cg}\) in \(\unit{tDM\,\unit{ha}^{-1}}\) and below-ground residues to above-ground biomass \(r^{\mathrm{bg}}_{cg}\) in \(\unit{tDM\,\unit{tDM}^{-1}}\) as follows

@@

Following the IPCC guidelines, we split the above-ground residue calculations into a yield-dependent slope (\(r^{\mathrm{ag,prod}}\)) and a positive intercept (\(r^{\mathrm{ag,area}}\)) fraction \citep{hergoualch_n2o_in_ipcc_2019}. Residues biomass therefore increases under-proportionally with rising yields, reflecting a shifting harvest index of higher-yielding breeds.
Deviating from Bodirsky et al.~\citeyearpar{bodirsky_n2o_2012} we use harvested instead of physical crop area (denoted in Eq. \eqref{eq:resbiomass} by MCF described in Sect. \ref{sec:landuse}) to account for increased residue biomass due to multiple cropping (multiple harvests with each lower yields) and decreased residue amounts due to fallow land. We assume that all \(BGR\) are left in the soil, whereas \(AGR\) can be burned or harvested for other purposes such as feeding animals \citep{weindl_livestock_2017}, fuel or for material use.

@@

From our 17 crop groups, we compile four residue groups (straw, high- and low-lignin residues, residues without dual use), of which the first three are taken away from the field for other purposes (see mappingCrop2Residue.csv in Bodirsky et al.~\citeyearpar{bodirsky_mrcommons_2020}). Residue feed demand for five different livestock groups is based on country-specific feed baskets \citep[see][]{weindl_livestock_2017}, that differentiate between the residue groups and take available \(AGR\) biomass as well as livestock productivity into account. We estimate a material-use share for the straw-residue group of 5\% and a fuel-share of 10\% for all used residue groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumed, and use shares of middle-income countries are linearly interpolated based on per-capita income, following the same rationale as for the share of burnt residues described above. The remaining \(AGR\) as well as all \(BGR\) are expected to be left on the field. We limit high residue return rates to at most \(10\,\unit{tC\,ha}^{-1}\) in order to correct for outliers.

@@

To disaggregate country level FAOSTAT livestock production data to half-degree resolution, we use the following rule-based assumptions, drawing from the approach of Robinson et al.~\citeyearpar{robinson_mapping_2014} and applying feed basket assumptions based on a revised methodology from Weindl et al.~\citeyearpar{weindl_livestock_2017}. We differentiate between ruminant and monogastric systems, as well as extensive and intensive systems.
Due to high feed demand of ruminants, we assume that ruminant livestock is located where the production of feed occurs to minimize transport of feed. We distinguish between grazed pasture, which is converted into livestock products in extensive systems, and primary-crop feed stuff, which we consider to be consumed in intensive systems.
For poultry, egg and monogastric meat production we use the per-capita income of the country to distinguish between intensive and extensive production systems. For low-income countries, we assume only extensive production systems. We locate them according to the share of built-up areas based on the assumption that these animals are held in subsistence or small-holder farming systems with a high labor-per-animal ratio. Intensive production associated with high-income countries, is distributed within a country using the share of primary-crop production, assuming that feed availability is the most determining factor for livestock location. For middle-income countries we split the livestock production into extensive and intensive systems based on the per-capita income.

Manure production and management is based on a revised methodology of Bodirsky et al.~\citeyearpar{bodirsky_n2o_2012} and is presented here due to its central role in soil carbon modeling. Based on the gridded livestock distribution we calculate spatially explicit excretion by estimating the nitrogen balance of the livestock systems on the basis of comprehensive livestock feed baskets \citep{weindl_livestock_2017}, assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed C:N ratios, which range from 10 for poultry up to 19 for beef cattle (for full detail see Calvo Buendia et al.~\citeyearpar{ogle_cropland_in_ipcc_2019}.
Depending on the feed system we assume manure to be handled in four different ways:
All manure originated from pasture feed intake is excreted directly on pastures and rangelands (pasture grazing), deducting manure collected as fuel.
Whereas for low-income countries, we adopt a share of 25\% of crop residues in feed intake directly consumed and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items, we assume the manure to be stored in animal waste management systems associated with livestock housing.
To estimate the carbon actually returned to the soil, we account for carbon losses during storage, where return shares depend on different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not returned to the fields. For manure stored in different animal waste management systems we compiled carbon loss rates (see calcClossConfinement.R in Bodirsky et al.~\citeyearpar{bodirsky_mrcommons_2020} for more details) depending on the different systems and the associated nitrogen loss rates as specified in Bodirsky et al.~\citeyearpar{bodirsky_n2o_2012}. We limit high application shares at \(10\,\unit{tC\,ha}^{-1}\) to correct for outliers, that can occur due to inconsistencies between FAO production and 0.5 degree land-use data.

@@

In order to derive a spatial distribution of the three different tillage types specified by the IPCC --- full tillage, reduced tillage and no tillage ---, we assume that all natural land and pastures are not tilled, whereas annual crops are under full and perennials under reduced tillage per default. Furthermore, we assume no tillage in cropland cells specified as no tillage cell based on the historical global gridded tillage data set from Porwollik et al.~\citeyearpar{porwollik_generating_2019}. This data set is extended to the period of 1975--2010 by combining country-level data on areas under conservation agriculture from FAO \citeyearpar{fao_aquastat_2016} and half-degree resolution physical crop areas from Hurtt et al.~\citeyearpar{hurtt_harmonization_2020}, applying the methodology of Prowollik et al.~\citeyearpar{porwollik_generating_2019} to identify potential no-tillage grid cells.

\hypertarget{sec:scenarios}{%
\subsection{Management scenario definitions}\label{sec:scenarios}}

To single out the impact of tillage practices, residue and manure inputs, we defined scenario with constant values for these three drivers: In the \textit{constTillage} scenario the adoption of no-tillage practices are neglected (adoption starts in 1974 according to the available data set). The \textit{constResidues} and the \textit{constManure} scenario assume constant input rates from residues resp. manure (in \(\unit{t ha^{-1}}\)) at the level of 1975 onward. Within the \textit{constResidue} scenario at different effects overlay each other: yields and with them residue biomass increase due to productivity gains; rates of residue left or returned to fields are raising; and shifts of cropping pattern change the amount of residue biomass due to crop-group specific harvest index values. The \textit{constManagement} combines all three scenarios \textit{constTillage}, \textit{constResidues} and \textit{constManure}.
\newpage

>>>>

\begin{abstract}
Soil organic carbon (SOC), one of the largest terrestrial carbon (C) stocks on Earth, has been depleted by anthropogenic land-cover change and agricultural management. However, the latter has so far not been well represented in global C stock assessments. While SOC models often simulate detailed biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated at the global scale. Here we develop a spatially explicit data set for agricultural management on cropland, considering crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We combine it with a reduced-complexity model based on the IPCC Tier 2 method to create a half-degree resolution data set of SOC stocks and SOC stock changes for the first 30 cm of mineral soils. We estimate that due to arable farming, soils have lost around \(34.6\,\unit{GtC}\) relative to a counterfactual hypothetical natural state in 1975. Within the period 1975--2010 this SOC debt continued to expand by \(5\,\unit{GtC}\) (\(0.14\,\unit{GtC\,yr^{-1}}\)) to around 39.6 GtC. However, accounting for historical management led to \(2.1\,\unit{GtC}\) less (\(0.06\,\unit{GtC\,yr^{-1}}\)) emissions than under the assumption of constant management. We also find that management decisions have influenced the historical SOC trajectory most strongly by residue returning, indicating that SOC enhancement by biomass retention may be a promising negative emissions technique. The reduced-complexity SOC model may allow to simulate management-induced SOC enhancement also within computationally demanding integrated (land-use) assessment modeling.
\end{abstract}

@@

Soil organic carbon (SOC), the amount of organic carbon stored in the Earth's soil, exceeds the carbon in the atmospheric and vegetation pools multiple times \citep{batjes_total_1996}. Even small changes in processes affecting SOC lead therefore to substantial shifts in the terrestrial carbon cycle and influence the amount of CO\textsubscript{2} in the atmosphere \citep{frielingstein_glob_carbon_2020, minasny_soil_2017}. The specific amount of carbon stored in soils globally is quantified with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile \citep{batjes_total_1996, sanderman_soil_2017}.

@@

The anthropogenic impact can be measured by the SOC debt (also referred to as SOC component of land-use change emissions, see Pongratz et al. \citeyearpar{pongratz_terminolgy_2014}), which is the amount of organic carbon soils have lost under cultivation compared to a hypothetical potential natural vegetation state. Sanderman et al. \citeyearpar{sanderman_soil_2017} identified the anthropogenic SOC debt for the first meter of the soil profile due to land cover change at around 116 GtC (37 GtC for the first 30 cm), compared to previous estimates of 60--130 GtC for the first meter \citep{lal_world_2001}.

Global assessments of the carbon cycle via dynamic global vegetation models (DGVMs), Earth System Models (ESMs) or bookkeeping models (BKMs) have analyzed SOC losses as part of a comprehensive evaluation of the global carbon budget and land-use change (LUC) emissions \citep{frielingstein_glob_carbon_2020}. While providing estimates of the magnitude of SOC losses due to land-cover change, most models lack a detailed representation of agricultural management. Earlier DGVM- and ESM-based assessments only considered changes in land cover, but ignored the removal of biomass at harvest \citep{strassmann_simulating_2008, betts_climate_2015}. BKMs are designed to estimate LUC related emissions and have largely improved in estimating additional emissions from wood harvest and shifting cultivation. However, state of the art models do not consider impacts of varying agricultural management \citep{frielingstein_glob_carbon_2020, houghton_carbon_2012, hansis_relevance_2015, bastos_comparison_2021}.

Managed agricultural systems were introduced in greater detail to DGVMs and ESMs to improve the assessment of the terrestrial carbon balance \citep[e.g.][]{bondeau_modelling_2007, lindeskog_implications_2013}. Pugh et al. \citeyearpar{pugh_simulated_2015} explicitly consider agricultural management in the form of tillage, irrigation and biomass extraction at harvest, but worked with uniform scenario assumptions on management rather than with historical management data. They also showed the importance of accounting for the land-use history, as many carbon emissions from agricultural soils are caused by historical LUC and the slow decline of SOC under cropland before a new equilibrium is reached.

In global-scale carbon cycle assessments, management systems are typically represented as spatially explicit patterns that are static in time \citetext{\citealp[e.g.~for growing seasons in][]{portmann_mirca2000global_2010}; \citealp[multiple cropping systems in][]{waha_multiple_2020}; \citealp[irrigation systems in][]{jagermeyr_water_2015}} or as stylized (in the sense of uniform management assumptions) scenarios \citep[e.g.][]{pugh_simulated_2015, lutz_simulating_2019}. Herzfeld et al. \citeyearpar{herzfeld_soc_2021} account for historical changes in fertilizer and manure inputs, residue removal rates and tillage systems and report SOC losses from cropland expansion over the period from 1700--2018 of 215 GtC. Within their stylized future management scenarios under future climate change they find that none of the management aspects considered (residue management, no-tillage) can create a positive SOC stock change on current cropland areas that counteracts the still continuing legacy flux from the initial land-use change.

More data sets on spatially explicit agricultural management time series with global coverage are becoming available (e.g.~on tillage systems, see \citep{porwollik_generating_2019, prestele_spatially_2018}) and modeling approaches are increasingly being developed to project the dynamics of management systems into the future (e.g. \citep{iizumi_modeling_2019, minoli_modelling_2019}), but have --- to our knowledge --- not yet found their way into comprehensive assessments of the terrestrial carbon cycle in DGVMs and BKMs.

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks \(SOC^{\mathrm{eq}}\) are given by
\begin{equation}
 SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
\label{eq:inoutflow}
\end{equation}
with \(C^{\textrm{in}}\) being the carbon inputs to the soil, \(k\) denotes the soil organic carbon decay rate. This equation is valid for all grid cells \(i\) and all years \(t\). We use the Tier 2 modeling approach for our calculations, which assumes three soil carbon sub-pools \(sub\) (active, slow and passive) and interactions between them, following the approach in the Century model \citep{parton_analysis_1987}. Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type \(lu\) specific.
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and grasslands (referred to as natural vegetation in the following). Forage crops are included within cropland, whereas pastures (including mowed meadows (perennials) and rangelands) are assigned to natural vegetation. Carbon flows connected to livestock are only considered in this study when they originate from cropland feed sources, while the manure originating from pasture biomass is disregarded, implicitly assuming that this manure is excreted or applied to pastures.

@@

Whereas residue and manure default lignin and nitrogen fractions are given by the IPCC guidelines \citep{ogle_cropland_in_ipcc_2019}, we use plant-functional type and plant-organ specific parameterization for the natural litterfall. Global distribution of plant-functional types is given by \citep{schaphoff_lpjml4_2018} as well as separation of litter into leaf, fine root and wood litter compartments excluding litter biomass burnt in wild fires. Leaf litter parameters are given by Brovkin et al. \citeyearpar{brovkin_plant-driven_2012}, fine root to leaf litter lignin ratio by Guo et al. \citeyearpar{guo_leaf_root_decompostion_2021}, lignin content of wood litter by Rahman et al. \citeyearpar{rahman_lignin_2013} and nitrogen content scaling factors for leaf to fine roots and leaf to wood litter by von Bloh et al. \citeyearpar{vBloh_lpjml5_2018}.
Data sources for all considered carbon inputs as well as for lignin and nitrogen parameterization are listed in Table \ref{tab:datasourceinputs}.

@@

For natural vegetation, we assume rainfed and non-tilled conditions, whereas for cropland, we distinguish the effect of different tillage (see Sect. \ref{sec:tillage}) and irrigation (see Sect. \ref{sec:irrigation}) practices on decay rates. We calculate area-weighted means for \(till\) and \(wat\) on cropland for each grid cell, using area shares for the different tillage and irrigation practices. Data sources as well as used parameters for the different decay drivers for all land-use types are listed in Table \ref{tab:datasourcedecay}; equations are displayed by equations 5.0B--5.0F in Ogle et al. \citeyearpar{ogle_cropland_in_ipcc_2019}.

@@

Firstly, in order to account for the impacts of legacy fluxes from land-use changes long before the time horizon of interest, we consider land-use change from 1510 onwards. In 1510, we assume all \(SOC\) pools to be in natural steady-state, implying that all land-use change prior to that time occurs in 1510. We assume that by 1901, all cropland converted in 1510 has reached its new steady state, so that it is not necessary to explicitly account for even older land conversion. Model inputs for 1901--1930 for climate and natural vegetation litterfall are repeated for 1510--1900 to mimic constant climate conditions for this first initializing period. Similarly, agricultural management data are held constant at the level of 1965 until 1965.
We acknowledge that this introduces a bias as agricultural management has changed prior to 1965, but this approach follows others studies on effects of land-use change and management \citep[e.g.][]{schaphoff_lpjml4_2018-1, herzfeld_soc_2021} and is limited by data availability on harvest statistics (and other management effects).

@@

Additionally to the Tier 2 modeling approach \citep{ogle_cropland_in_ipcc_2019} and the detailed analysis of management data coming with it, \(SOC\) changes can be estimated using the IPCC Tier 1 approach of IPCC guidelines \citep{eggleston_ipcc_2006, calvo_buendia_ipcc_2019}. Here, stocks are calculated via stock change factors (\(F^{\mathrm{SCF}}\)) given by the IPCC for the topsoil (0-30 cm) and based on observational data. Note that IPCC factors are derived under the assumption that there is a linear change between steady states over 20 years. Estimates of \(F^{\mathrm{SCF}}\) are differentiated by crops, management and input systems (here summarized under \(m\)) reflecting different dynamics under changed in- and outflows without explicitly tracking these flows. Moreover, estimates of \(F^{\mathrm{SCF}}\) vary for different climatic zones (\(c\)) specified by the IPCC (see Fig. \ref{fig:CLIMzone}). The actual \(SOC\) stocks are thus calculated based on a given reference stock \(SOC^{\mathrm{ref}}\) by

@@

with \(T_{c,i}\) being the translation matrix for grid cells \(i\) into corresponding climate zones \(c\). For this analysis, we use the default \(F^{\mathrm{SCF}}\) from the Tier 1 method of Eggleston et al. \citeyearpar{eggleston_ipcc_2006} and Calvo Buendia et al. \citeyearpar{calvo_buendia_ipcc_2019} as a comparison and consistency check for our more detailed Tier 2 steady-state approach.

@@

Crop residue production and management is based on a revised methodology of Bodirsky et al. \citeyearpar{bodirsky_n2o_2012} and key aspects are explained as they play a central role in soil carbon modeling. Starting from crop yield estimates \(Y\) and respective physical crop area \(CA\), we estimate total above-ground \(AGR\) and below-ground \(BGR\) residue biomass (in tonnes) using crop group \(cg\) specific ratios for above-ground residues to harvested biomass \(r^{\mathrm{ag,prod}}_{cg}\) in \((tDM\,ha^{-1})(tDM\,ha^{-1})^{-1}\) above-ground residues to harvested area \(r^{\mathrm{ag,area}}_{cg}\) in \(\unit{tDM\,\unit{ha}^{-1}}\) and below-ground residues to above-ground biomass \(r^{\mathrm{bg}}_{cg}\) in \(\unit{tDM\,\unit{tDM}^{-1}}\) as follows

@@

Following the IPCC guidelines, we split the above-ground residue calculations into a yield-dependent slope (\(r^{\mathrm{ag,prod}}\)) and a positive intercept (\(r^{\mathrm{ag,area}}\)) fraction \citep{hergoualch_n2o_in_ipcc_2019}. Residues biomass therefore increases under-proportionally with rising yields, reflecting a shifting harvest index of higher-yielding breeds.
Deviating from Bodirsky et al. \citeyearpar{bodirsky_n2o_2012} we use harvested instead of physical crop area (denoted in Eq. \eqref{eq:resbiomass} by MCF described in Sect. \ref{sec:landuse}) to account for increased residue biomass due to multiple cropping (multiple harvests with each lower yields) and decreased residue amounts due to fallow land. We assume that all \(BGR\) are left in the soil, whereas \(AGR\) can be burned or harvested for other purposes such as feeding animals \citep{weindl_livestock_2017}, fuel or for material use.

@@

From our 17 crop groups, we compile four residue groups (straw, high- and low-lignin residues, residues without dual use), of which the first three are taken away from the field for other purposes (see mappingCrop2Residue.csv in Bodirsky et al. \citeyearpar{bodirsky_mrcommons_2020}). Residue feed demand for five different livestock groups is based on country-specific feed baskets \citep[see][]{weindl_livestock_2017}, that differentiate between the residue groups and take available \(AGR\) biomass as well as livestock productivity into account. We estimate a material-use share for the straw-residue group of 5\% and a fuel-share of 10\% for all used residue groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumed, and use shares of middle-income countries are linearly interpolated based on per-capita income, following the same rationale as for the share of burnt residues described above. The remaining \(AGR\) as well as all \(BGR\) are expected to be left on the field. We limit high residue return rates to at most \(10\,\unit{tC\,ha}^{-1}\) in order to correct for outliers.

@@

To disaggregate country level FAOSTAT livestock production data to half-degree resolution, we use the following rule-based assumptions, drawing from the approach of Robinson et al. \citeyearpar{robinson_mapping_2014} and applying feed basket assumptions based on a revised methodology from Weindl et al. \citeyearpar{weindl_livestock_2017}. We differentiate between ruminant and monogastric systems, as well as extensive and intensive systems.
Due to high feed demand of ruminants, we assume that ruminant livestock is located where the production of feed occurs to minimize transport of feed. We distinguish between grazed pasture, which is converted into livestock products in extensive systems, and primary-crop feed stuff, which we consider to be consumed in intensive systems.
For poultry, egg and monogastric meat production we use the per-capita income of the country to distinguish between intensive and extensive production systems. For low-income countries, we assume only extensive production systems. We locate them according to the share of built-up areas based on the assumption that these animals are held in subsistence or small-holder farming systems with a high labor-per-animal ratio. Intensive production associated with high-income countries, is distributed within a country using the share of primary-crop production, assuming that feed availability is the most determining factor for livestock location. For middle-income countries we split the livestock production into extensive and intensive systems based on the per-capita income.

Manure production and management is based on a revised methodology of Bodirsky et al. \citeyearpar{bodirsky_n2o_2012} and is presented here due to its central role in soil carbon modeling. Based on the gridded livestock distribution we calculate spatially explicit excretion by estimating the nitrogen balance of the livestock systems on the basis of comprehensive livestock feed baskets \citep{weindl_livestock_2017}, assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed C:N ratios, which range from 10 for poultry up to 19 for beef cattle (for full detail see Calvo Buendia et al. \citeyearpar{ogle_cropland_in_ipcc_2019}.
Depending on the feed system we assume manure to be handled in four different ways:
All manure originated from pasture feed intake is excreted directly on pastures and rangelands (pasture grazing), deducting manure collected as fuel.
Whereas for low-income countries, we adopt a share of 25\% of crop residues in feed intake directly consumed and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items, we assume the manure to be stored in animal waste management systems associated with livestock housing.
To estimate the carbon actually returned to the soil, we account for carbon losses during storage, where return shares depend on different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not returned to the fields. For manure stored in different animal waste management systems we compiled carbon loss rates (see calcClossConfinement.R in Bodirsky et al. \citeyearpar{bodirsky_mrcommons_2020} for more details) depending on the different systems and the associated nitrogen loss rates as specified in Bodirsky et al. \citeyearpar{bodirsky_n2o_2012}. We limit high application shares at \(10\,\unit{tC\,ha}^{-1}\) to correct for outliers, that can occur due to inconsistencies between FAO production and 0.5 degree land-use data.

@@

In order to derive a spatial distribution of the three different tillage types specified by the IPCC --- full tillage, reduced tillage and no tillage ---, we assume that all natural land and pastures are not tilled, whereas annual crops are under full and perennials under reduced tillage per default. Furthermore, we assume no tillage in cropland cells specified as no tillage cell based on the historical global gridded tillage data set from Porwollik et al. \citeyearpar{porwollik_generating_2019}. This data set is extended to the period of 1975--2010 by combining country-level data on areas under conservation agriculture from FAO \citeyearpar{fao_aquastat_2016} and half-degree resolution physical crop areas from Hurtt et al. \citeyearpar{hurtt_harmonization_2020}, applying the methodology of Prowollik et al. \citeyearpar{porwollik_generating_2019} to identify potential no-tillage grid cells.

\hypertarget{sec:scenarios}{%
\subsection{Scenario definitions}\label{sec:scenarios}}

To highlight the impact of changing management effects and to assess the sensitivity of the model towards different initialization choices, we perform a set of scenario runs. In the following section we outline name and idea of these scenarios \citep[for technical implementation see][]{karstens_mrsoil_2020}.

To single out the impact of tillage practices, residue and manure inputs, we defined scenario with constant values for these three drivers: In the \textit{constTillage} scenario the adoption of no-tillage practices are neglected (adoption starts in 1974 according to the available data set). The \textit{constResidues} and the \textit{constManure} scenario assume constant input rates from residues resp. manure (in \(\unit{t ha^{-1}}\)) at the level of 1975 onward. Within the \textit{constResidue} scenario at different effects overlay each other: yields and with them residue biomass increase due to productivity gains; rates of residue left or returned to fields are raising; and shifts of cropping pattern change the amount of residue biomass due to crop-group specific harvest index values. The \textit{constManagement} combines all three scenarios \textit{constTillage}, \textit{constResidues} and \textit{constManure}.
